---
title: "Reproduction of Simulation 1 by Jamieson, Crump & Hannah (2012)"
author: "Frederik Aust"
date: "13.1.2015"
output:
  html_document:
    theme: spacelab
    code_folding: show
    toc: yes
    toc_float: yes
---

To validate this implementation of Minerva-AL, I reproduced a simulation study reported in Jamieson, Crump & Hannah (2012).

```{r echo = FALSE}
library("dplyr")
library("tidyr")

# Run fortran simulation
if(.Platform$OS.type == "unix") {
  system("cd reference_implementation; sh make.sh")
}

source("minerva-al.R")
```

<br />

# Simulation of acquisition and extinction (Section 1)

I created cue, outcome and context vectors and defined the number of trials and replications according to the specifications in the paper.

```{r setup-events}
n_features <- 120
cue_features <- 1:100
a <- context <- outcome <- rep(0, n_features)

a[1:20] <- 1
outcome[101:120] <- 1
context[81:100] <- 1

acquisition_event <- a + context + outcome
extinction_event <- probe <- a + context
```

```{r setup-simulation}
n_replications <- 100
n_trials <- 200

p_encode <- c(1/3, 2/3, 1)
```

```{r simulate-acquisition}
sim_results <- matrix(0, ncol = n_trials, nrow = length(p_encode))

for (r in 1:n_replications) {
  for (i in 1:3) {
    # Memory is empty on first trial
    normalized_echo <- probe_memory(probe, NULL, cue_features)
    expectancy <- expect_event(outcome, normalized_echo)
    memory <- learn(
      normalized_echo
      , acquisition_event
      , p_encode[i]
      , NULL
    )
    
     # Acquisition trials
    for(j in 2:(n_trials / 2)) {
      normalized_echo <- probe_memory(probe, memory, cue_features)
      expectancy <- c(expectancy, expect_event(outcome, normalized_echo))
      memory <- learn(
        normalized_echo
        , acquisition_event
        , p_encode[i]
        , memory
      )
    }
    
    # Extinction trials
    for(j in ((n_trials / 2) + 1):n_trials) {
      normalized_echo <- probe_memory(probe, memory, cue_features)
      expectancy <- c(expectancy, expect_event(outcome, normalized_echo))
      memory <- learn(
        normalized_echo
        , extinction_event
        , p_encode[i]
        , memory
      )
    }
    
    sim_results[i, ] <- sim_results[i, ] + expectancy
  }
}

sim_results <- sim_results / n_replications # Mean of replications
```

<br />

## Results

The resulting expectancies correspond nicely to those reported by Jamieson, Hannah & Crump (2012) in Figure 1.

```{r plot-acquisition, fig.height = 5.5, fig.width = 9.5, echo = FALSE}
plot(
  1:200
  , rep(NA, 200)
  , ylim = c(0, 1)
  , lwd = 2.5
  , xlab = "Trial"
  , ylab = "Retrieval of X given A"
  , las = 1
)

matlines(
  t(sim_results)
  , lwd = 2.5
  , col = "black"
  , lty = 1
)

matpoints(
  t(sim_results)
  , pch = c(24, 22, 21)
  , bg = "white"
  , col = "black"
  , cex = 1.25
)
```

As a comparison, the following plot shows the results from the reference implementation I received from Randall Jamieson (thanks!).

```{r}
acquisition_files <- list.files(path = "reference_implementation/results", pattern = "Acquisition", full.names = TRUE)

acquisition_extinction <- lapply(
  acquisition_files
  , read.delim
  , skip = 3
  , nrows = 24
  , sep = ""
  , header = FALSE
) %>% 
  setNames(
    nm = stringr::str_extract(basename(acquisition_files), "\\d+") %>%
      gsub("0", "0.", .)
  ) %>% 
  bind_rows(.id = "L") %>% 
  mutate(V2 = as.factor(V2)) %>% 
  group_by(L) %>% 
  summarise_if(is.numeric, mean)
```

```{r plot-acquisition2, fig.height = 5.5, fig.width = 9.5, echo = FALSE}
plot(
  1:200
  , rep(NA, 200)
  , ylim = c(0, 1)
  , lwd = 2.5
  , xlab = "Trial"
  , ylab = "Retrieval of X given A"
  , las = 1
)

matlines(
  t(as.matrix(acquisition_extinction[, -1]))
  , lwd = 2.5
  , col = "black"
  , lty = 1
)

matpoints(
  t(as.matrix(acquisition_extinction[, -1]))
  , pch = c(24, 22, 21)
  , bg = "white"
  , col = "black"
  , cex = 1.25
)
```

```{r}
prediction_differences <- sim_results - as.matrix(acquisition_extinction[, -1])

plot(
  1:200
  , rep(NA, 200)
  , ylim = c(-0.2, 0.2)
  , lwd = 2.5
  , xlab = "Trial"
  , ylab = "Absolut difference"
  , las = 1
)

matlines(
  t(prediction_differences)
  , col = "black"
  , lty = c(1, 3, 5)
)

legend("topright", inset = 0.05, legend = unlist(acquisition_extinction[, 1]), lty = c(1, 3, 5))

summary(as.vector(prediction_differences))
```

Additionally, the following plots visualize the information encoded for cue and outcome features across acquisition and extinction trials. The data are taken from one of the `r n_replications` simulations with encoding probability $L = 1$.

```{r plot-encoding, echo = FALSE}
plot(
  1:200
  , memory[, 1]
  , type = "l"
  , col = scales::alpha("black", 0.3)
  , ylim = c(-2, 2)
  , xlab = "Trial"
  , ylab = "Feature encoding"
  , main = "Features of cue A"
  , las = 1
)
for(i in 2:20) {
  lines(
    1:200
    , memory[, i]
    , col = scales::alpha("black", 0.3)
  )
}


plot(
  1:200
  , memory[, 101]
  , type = "l"
  , col = scales::alpha("black", 0.3)
  , ylim = c(-2, 2)
  , xlab = "Trial"
  , ylab = "Feature encoding"
  , main = "Features of outcome X"
  , las = 1
)
for(i in 102:120) {
  lines(
    1:200
    , memory[, i]
    , col = scales::alpha("black", 0.3)
  )
}
```


# Simulation of reacquisition (Section 1)

```{r}
b <- rep(0, n_features)

b[21:40] <- 1
control_event <- b + context + outcome
control_probe <- b + context
```

```{r}
n_replications <- 100
n_trials <- 200
```


```{r simulate-reacquisition}
reacquisition_sim_results <- control_sim_results <- matrix(NA, ncol = n_trials, nrow = length(p_encode) * n_replications)

for (i in 1:length(p_encode)) {
  for (r in 1:n_replications) {
    # Memory is empty on first trial
    normalized_echo <- probe_memory(probe, NULL, cue_features)
    expectancy <- expect_event(outcome, normalized_echo)
    memory <- learn(
      normalized_echo
      , acquisition_event
      , p_encode[i]
      , NULL
    )
    
    # Acquisition trials
    for(j in 2:50) {
      normalized_echo <- probe_memory(probe, memory, cue_features)
      expectancy <- c(expectancy, expect_event(outcome, normalized_echo))
      memory <- learn(
        normalized_echo
        , acquisition_event
        , p_encode[i]
        , memory
      )
    }
    
    # Extinction trials
    for(j in 51:100) {
      normalized_echo <- probe_memory(probe, memory, cue_features)
      expectancy <- c(expectancy, expect_event(outcome, normalized_echo))
      memory <- learn(
        normalized_echo
        , extinction_event
        , p_encode[i]
        , memory
      )
    }
    
    # Reacquisition trials
    reacquisition_memory <- memory
    reacquisition_expectancy <- expectancy
    
    for(j in 101:200) {
      normalized_echo <- probe_memory(probe, reacquisition_memory, cue_features)
      reacquisition_expectancy <- c(reacquisition_expectancy, expect_event(outcome, normalized_echo))
      reacquisition_memory <- learn(
        normalized_echo
        , acquisition_event
        , p_encode[i]
        , reacquisition_memory
      )
    }
    
    # Control trials
    control_memory <- memory
    control_expectancy <- expectancy
    
    for(j in 101:200) {
      normalized_echo <- probe_memory(control_probe, control_memory, cue_features)
      control_expectancy <- c(control_expectancy, expect_event(outcome, normalized_echo))
      control_memory <- learn(
        normalized_echo
        , control_event
        , p_encode[i]
        , control_memory
      )
    }
    
    reacquisition_sim_results[(i-1)*n_replications + r, ] <- reacquisition_expectancy
    control_sim_results[(i-1)*n_replications + r, ] <- control_expectancy
  }
}
```

## Results

```{r fig.height = 5.5, fig.width = 9.5, echo = FALSE}
plot(
  51:200
  , rep(0, 150)
  , pch = NA
  , ylim = c(0, 1)
  , lwd = 2.5
  , xlab = "Trial"
  , ylab = "Retrieval of X given cue"
  , las = 1
  , 
)

abline(h = 0.95, col = "grey")

for(i in 1:length(p_encode)) {
  lines(51:200, colMeans(reacquisition_sim_results[(1:n_replications) + (i-1)*n_replications, 51:200]), lwd = 2)
  lines(51:200, colMeans(control_sim_results[(1:n_replications) + (i-1)*n_replications, 51:200]), lty = "dashed")
}


mean_se <- function(x) paste0(round(mean(x), 2), " (", round(sd(x) / sqrt(length(x)), 2), ")")

reacquisition_results <- matrix(NA, ncol = 3, nrow = 2)

for(i in 1:length(p_encode)) {
  reacquisition_results[, i] <- c(
    mean_se(apply(reacquisition_sim_results[(1:n_replications) + (i-1)*n_replications, 101:200], 1, function(x) min(which(x >= 0.95))))
    , mean_se(apply(control_sim_results[(1:n_replications) + (i-1)*n_replications, 101:200], 1, function(x) min(which(x >= 0.95))))
  )
}

knitr::kable(cbind(Condition = c("Reacquisition", "Control"), reacquisition_results), col.names = c("Condtion", round(p_encode, 2)))
```

As a comparison, the following plot shows the results from the reference implementation I received from Randall Jamieson (thanks!).

```{r}
reacquisition_files <- list.files(path = "reference_implementation/results", pattern = "Reacquisition\\_\\d+", full.names = TRUE)

reacquisition <- lapply(
  reacquisition_files
  , read.delim
  , skip = 3
  , nrows = 24
  , sep = ""
  , header = FALSE
) %>% 
  setNames(
    nm = stringr::str_extract(basename(reacquisition_files), "\\d+") %>%
      gsub("0", "0.", .)
  ) %>% 
  bind_rows(.id = "L") %>% 
  mutate(V2 = as.factor(V2))

mean_reacquisition <- reacquisition %>% 
  group_by(L) %>% 
  summarise_if(is.numeric, mean)
```

```{r}
reacquisition_control_files <- list.files(path = "reference_implementation/results", pattern = "Reacquisition\\_\\D+", full.names = TRUE)

reacquisition_control <- lapply(
  reacquisition_control_files
  , read.delim
  , skip = 3
  , nrows = 24
  , sep = ""
  , header = FALSE
) %>% 
  setNames(
    nm = stringr::str_extract(basename(reacquisition_control_files), "\\d+") %>%
      gsub("0", "0.", .)
  ) %>% 
  bind_rows(.id = "L") %>% 
  mutate(V2 = as.factor(V2))
  
mean_reacquisition_control <- reacquisition_control %>% 
  group_by(L) %>% 
  summarise_if(is.numeric, mean)
```

```{r plot-reacquisition, fig.height = 5.5, fig.width = 7.5, echo = FALSE}
plot(
  1:150
  , rep(NA, 150)
  , ylim = c(0, 1)
  , lwd = 2.5
  , xlab = "Trial"
  , ylab = "Retrieval of X given A"
  , las = 1
)

abline(h = 0.95, col = "grey")

matlines(
  cbind(51:200, t(as.matrix(mean_reacquisition[, 52:201])))
  , col = "black"
  , lty = 1
  , lwd = 2
)

matlines(
  cbind(51:200, t(as.matrix(mean_reacquisition_control[, 52:201])))
  , col = "black"
  , lty = 2
)
```

```{r}
reacquisition_results <- matrix(NA, ncol = 3, nrow = 2)

trails_to_master <- function(x) data.frame(n_trials = min(which(x >= 0.95)))

reacquisition_trial_counts <- reacquisition %>%
  group_by(L, V2) %>% 
  do(trails_to_master(.[, 104:ncol(reacquisition)])) %>% 
  group_by(L) %>% 
  summarize(n_trials = mean_se(n_trials))

reacquisition_control_trial_counts <- reacquisition_control %>%
  group_by(L, V2) %>% 
  do(trails_to_master(.[, 104:ncol(reacquisition_control)])) %>% 
  group_by(L) %>% 
  summarize(n_trials = mean_se(n_trials))

knitr::kable(
  bind_rows(Reacquition = reacquisition_trial_counts, Control = reacquisition_control_trial_counts, .id = "Condition") %>%
    spread(L, n_trials) %>% 
    arrange(desc(Condition))
)
```

The following plot compares initial and reacquisition.

```{r}
plot(
  1:50
  , rep(0, 50)
  , pch = NA
  , ylim = c(0, 1)
  , lwd = 2.5
  , xlab = "Trial"
  , ylab = "Retrieval of X given A"
  , las = 1
)

abline(h = 0.95, col = "grey")

for(i in 1:length(p_encode)) {
  lines(colMeans(reacquisition_sim_results[(1:n_replications) + (i-1)*n_replications, 1:50]), lty = 2)
}

for(i in 1:length(p_encode)) {
  lines(colMeans(reacquisition_sim_results[(1:n_replications) + (i-1)*n_replications, 101:150]), lwd = 2)
}
```

And the same for the reference implementation.

```{r}
plot(
  1:50
  , rep(NA, 50)
  , ylim = c(0, 1)
  , lwd = 2.5
  , xlab = "Trial"
  , ylab = "Retrieval of X given A"
  , las = 1
)

abline(h = 0.95, col = "grey")

matlines(
  cbind(1:50, t(as.matrix(mean_reacquisition[, 102:151])))
  , col = "black"
  , lty = 1
  , lwd = 2
)

matlines(
  cbind(1:50, t(as.matrix(mean_reacquisition[, 2:51])))
  , col = "black"
  , lty = 2
)
```


<br />

# References

Jamieson, R. K., Crump, M. J. C., & Hannah, S. D. (2012). An instance theory of associative learning. *Learning & Behavior*, 40(1), 61–82. doi:[10.3758/s13420-011-0046-2](http://dx.doi.org/10.3758/s13420-011-0046-2)
