---
title: "Reproduction of Simulation 1 by Jamieson, Crump & Hannah (2012)"
author: "Frederik Aust"
date: "13.1.2015"
output:
  html_document:
    theme: spacelab
    toc: yes
---

To validate this implementation of Minerva-AL, I reproduced a simulation study reported in Jamieson, Crump & Hannah (2012).

```{r echo = FALSE}
source("minerva-al.R")
```

<br />

# Simulation of acquisition and extinction (Section 1)
I created cue, outcome and context vectors and defined the number of trials and replications according to the specifications in the paper.

```{r setup_events}
n_features <- 120
cue_features <- 1:100
a <- context <- outcome <- rep(0, n_features)

a[1:20] <- 1
outcome[101:120] <- 1
context[81:100] <- 1

acquisition_event <- a + context + outcome
extinction_event <- probe <- a + context
```

```{r setup_simulation}
n_replications <- 25
n_trials <- 200

p_encode <- c(0.33, 0.67, 1)
```

```{r execute_simulation}
sim_results <- matrix(0, ncol = n_trials, nrow = length(p_encode))

for (r in 1:n_replications) {
  for(i in 1:3) {
    # Memory is empty on first trial
    normalized_echo <- calc_echo(probe, NULL, cue_features)
    expectancy <- calc_expectancy(outcome, normalized_echo)
    memory <- learn(
      normalized_echo
      , acquisition_event
      , p_encode[i]
      , NULL
    )
    
     # Acquisition trials
    for(j in 2:(n_trials / 2)) {
      normalized_echo <- calc_echo(probe, memory, cue_features)
      expectancy <- c(expectancy, calc_expectancy(outcome, normalized_echo))
      memory <- learn(
        normalized_echo
        , acquisition_event
        , p_encode[i]
        , memory
      )
    }
    
    # Extinction trials
    for(j in ((n_trials / 2) + 1):n_trials) {
      normalized_echo <- calc_echo(probe, memory, cue_features)
      expectancy <- c(expectancy, calc_expectancy(outcome, normalized_echo))
      memory <- learn(
        normalized_echo
        , extinction_event
        , p_encode[i]
        , memory
      )
    }
    
    sim_results[i, ] <- sim_results[i, ] + expectancy
  }
}

sim_results <- sim_results / n_replications # Mean of replications
```

<br />

# Results
The resulting expectancies correspond nicely to those reported by Jamieson, Hannah & Crump (2012) in Figure 1.

```{r plot_results, fig.height = 5.5, fig.width = 9.5, echo = FALSE}
plot(
  1:n_trials, sim_results[1, ]
  , ylim = c(0, 1)
  , type = "l"
  , lwd = 2.5
  , xlab = "Trial"
  , ylab = "Retrieval of X given A"
  , las = 1
)
points(1:n_trials, sim_results[1, ], pch = 17, cex = 1.25, col = "white")
points(1:n_trials, sim_results[1, ], pch = 2, cex = 1.25)

lines(1:n_trials, sim_results[2, ], lwd = 2.5)
points(1:n_trials, sim_results[2, ], pch = 15, cex = 1.25, col = "white")
points(1:n_trials, sim_results[2, ], pch = 0, cex = 1.25)

lines(1:n_trials, sim_results[3, ], lwd = 2.5)
points(1:n_trials, sim_results[3, ], pch = 16, cex = 1.25, col = "white")
points(1:n_trials, sim_results[3, ], cex = 1.25)
```

Additionally, the following plots visualize the information encoded for cue and outcome features across acquisition and extinction trials. The data are taken from one of the `r n_replications` simulations with encoding probability $L = 1$.

```{r plot_encoding, echo = FALSE}
plot(
  1:200
  , memory[, 1]
  , type = "l"
  , col = scales::alpha("black", 0.3)
  , ylim = c(-2, 2)
  , xlab = "Trial"
  , ylab = "Feature encoding"
  , main = "Features of cue A"
  , las = 1
)
for(i in 2:20) {
  lines(
    1:200
    , memory[, i]
    , col = scales::alpha("black", 0.3)
  )
}


plot(
  1:200
  , memory[, 101]
  , type = "l"
  , col = scales::alpha("black", 0.3)
  , ylim = c(-2, 2)
  , xlab = "Trial"
  , ylab = "Feature encoding"
  , main = "Features of outcome X"
  , las = 1
)
for(i in 102:120) {
  lines(
    1:200
    , memory[, i]
    , col = scales::alpha("black", 0.3)
  )
}
```

<br />

# References
Jamieson, R. K., Crump, M. J. C., & Hannah, S. D. (2012). An instance theory of associative learning. *Learning & Behavior*, 40(1), 61–82. doi:[10.3758/s13420-011-0046-2](http://dx.doi.org/10.3758/s13420-011-0046-2)
