---
title: "Reproduction of simulations by Hintzman (1984)"
author: "Frederik Aust"
date: "06.02.2015"
output:
  html_document:
    theme: spacelab
    toc: yes
---

To validate this implementation of MINERVA2, I reproduced small parts of the simulations reported in Hintzman (1988).

```{r echo = FALSE}
source("minerva2.R")
```

<br />

# Frequency judgements
For reasons of simplicity, I simulated one subject with 5000 runs to yield sufficiently smooth intensity distribution plots (instead of 1000 subjects with 1000 runs each).

```{r setup_simulation}
frequencies <- 1:5

n_features <- 20
n_items <- 20

p_encode <- 0.5
```

```{r frequency_judgments, cache = TRUE}
results <- c()

for(run in 1:5000) {
  # Generate items
  item_features <- sample(c(-1, 0, 1), n_items * n_features, replace = TRUE)
  items <- matrix(item_features, ncol = n_features)
  
  control_features <- sample(c(-1, 0, 1), 4 * n_features, replace = TRUE)
  control_items <- matrix(control_features, ncol = n_features)
  
  item_frequencies <- rep(frequencies, each = 4)
  
  # Set up memory
  ## Save four items per level of frequency into memory with L = 0.5
  memory <- c()
  for(i in 1:n_items) {
    new_traces <- rep(items[i, ], item_frequencies[i])
    new_traces <- matrix(new_traces, ncol = n_features, byrow = TRUE)
    memory <- rbind(memory, new_traces)
  }
  memory <- forget(memory, p_encode)
  
  
  # Test memory
  intensities <- c()
  
  ## Control items (frequency = 0)
  control_intensity <- apply(control_items, 1, function(x) probe_memory(x, memory)$intensity)
  intensities <- cbind(intensities, control_intensity)
  
  ## Learned items (frequency = [1, 4])
  for(i in frequencies) {
    probes <- items[which(item_frequencies == i), ]
    
    intensity <- apply(probes, 1, function(x) probe_memory(x, memory)$intensity)
    intensities <- cbind(intensities, intensity)
  }
  
  results <- rbind(results, intensities)
}

colnames(results) <- paste0("freq.", c(0, frequencies))
```

## Results
The resulting echo intensities allow for a close reproduction of Hintzman's Figure 1 (1988).

```{r echo = FALSE}
plot(
  density(results[, 1])
  , xlim = c(-0.5, 1.5)
  , xlab = "Echo Intensity"
  , ylab = "Probability"
  , main = ""
  , axes = FALSE
  , lwd = 2
)
invisible(apply(results[, 2:6], 2, function(x) lines(density(x), lwd = 2)))
abline(h = 0, lwd = 2)
axis(1)
box()
```

<br />

# References
Hintzman, D. L. (1988). Judgments of frequency and recognition memory in a multiple-trace memory model. *Psychological Review*, 95(4), 528–551. doi:[10.1037/0033-295X.95.4.528](http://dx.doi.org/10.1037/0033-295X.95.4.528)
