---
title: "Reproduction of model-based analyses by Nosofsky (1989)"
author: "Frederik Aust"
date: "19.12.2014"
output:
  html_document:
    theme: spacelab
    toc: yes
---

To validate this implementation of the Generalized Context Model (GCM), I reproduced small parts of the model-based analyses reported in Nosofsky (1989). The original MDS solutions and response data were provided by Robert Nosofsky (s. [note on data](data/README.html)).

```{r echo = FALSE}
source("gcm_pred.r")
source("gcm_fit.r")
```

```{r}
sims <- read.csv2("data/nosofsky_1989_similarities.csv")[, -1]
sims$size <- c(0, 1, 0, 0, 1, 0, 1, 1, 0, 0, 2, 2, 0, 2, 0, 0)
sims$angle <- c(0, 1, 2, 0, 0, 1, 2, 0, 1, 0, 0, 2, 0, 0, 2, 0)
  
data <- read.csv2("data/nosofsky_1989_responses.csv")[, -1]
data$n_size <- rowSums(data[, 1:2])
data$n_angle <- rowSums(data[, 3:4])
```

<br />

# Unconstrained GCM fits for the size condition
The original parameter estimates reported in Table 5 by Nosofsky (1989) are $w_1 = .10$, $c = 1.60$, and $b_1 = .50$.

```{r}
obs <- sims[, 1:2]
mem <- subset(sims, size != 0)
size_fit <- optim(
  par = c(0.5, 2, 0.5)
  , mem = mem[, 1:3]
  , obs = obs
  , rho = 2
  , p = 2
  , data = data[, c(1, 2, 5)]
  , fn = gcm_fit
  , method = "Nelder-Mead"
)

size_fit$par

size_pred <- gcm_pred(size_fit$par, mem[, 1:3], obs, rho = 2, p = 2)
```

<br />

# Unconstrained GCM fits for the angle condition
The original parameter estimates reported in Table 5 by Nosofsky (1989) are $w_1 = .98$, $c = 3.20$, and $b_1 = .43$.

```{r}
mem <- subset(sims, angle != 0)
angle_fit <- optim(
  par = c(0.5, 2, 0.5)
  , mem = mem[, c(1:2, 4)]
  , obs = obs
  , rho = 2
  , p = 2
  , data = data[, c(3, 4, 6)]
  , fn = gcm_fit
  , method = "Nelder-Mead"
)

angle_fit$par

angle_pred <- gcm_pred(angle_fit$par, mem[, c(1:2, 4)], obs, rho = 2, p = 2)
```

<br />

# Predictions
The resulting fits allow for a close partial reproduction of Nosofsky's Figure 6 (1989) plotting observed against predicted proportions of category 1 responses for each stimulus.

```{r echo = FALSE}
par(pty = "s")
plot(
  data$Cat.1.s / data$n_size
  , size_pred
  , xlab = "Observed probability"
  , ylab = "Predicted probability"
  , pch = 17
  , asp = 1
  , las = 1
)
points(
  data$Cat.1.a / data$n_angle
  , angle_pred
  , pch = 0
)
abline(0, 1)
legend(
  "topleft"
  , legend = c("Size", "Angle")
  , pch = c(17, 0)
  , inset = 0.1
  , bty = "n"
)
```

<br />

# References
Nosofsky, R. M. (1989). Further tests of an exemplar-similarity approach to relating identification and categorization. *Perception & Psychophysics*, 45(4), 279–290. doi: [10.3758/BF03204942](http://dx.doi.org/10.3758/BF03204942)
