---
title: "Reproduction of model-based analyses by Nosofsky (1989)"
author: "Frederik Aust"
output:
  knitrBootstrap::bootstrap_document:
    highlight: xcode
    theme: flatly
    menu: false
---

To validate this implementation of the Bayesian Generalized Context Model (GCM), I reproduced small parts of the model-based analyses reported in Nosofsky (1989). The original MDS solutions and response data were provided by Robert Nosofsky (s. [note on data](../data/README.html)).

```{r echo = FALSE, message = FALSE}
library("runjags")
library("vioplot")
```

```{r}
similarities <- read.csv2("../data/nosofsky_1989_similarities.csv")[, -1]
similarities$size <- c(0, 1, 0, 0, 1, 0, 1, 1, 0, 0, 2, 2, 0, 2, 0, 0)
similarities$angle <- c(0, 1, 2, 0, 0, 1, 2, 0, 1, 0, 0, 2, 0, 0, 2, 0)
  
data <- read.csv2("../data/nosofsky_1989_responses.csv")[, -1]
data$n_size <- rowSums(data[, 1:2])
data$n_angle <- rowSums(data[, 3:4])
```

<br />

# Unconstrained GCM fits for the size condition
The original parameter estimates reported in Table 5 by Nosofsky (1989) are $w_1 = .10$, $c = 1.60$, and $b_1 = .50$.

```{r message = FALSE, results = "hide"}
tests <- as.matrix(similarities[, 1:2])
memory <- as.matrix(subset(similarities, size != 0)[, 1:2])
size_data <- list(
  y = data$Cat.1.s
  , tests = tests
  , memory = memory
  , ntrials = data$n_size
  , p = 2 # Shape of relationship between similarity and psychological distance
  , rho = 2 # Power of the Minkowski distance
  , category = unlist(subset(similarities, size != 0)[, 3])
)

init_values <- list(
  list(c = 4, w = 0.5, b = 0.5)
  , list(c = 2, w = 0.8, b = 0.3)
  , list(c = 3, w = 0.2, b = 0.7)
)

poi <- c(unique(unlist(lapply(init_values, names))), "pred_y")

# Sample
size_samples <- run.jags(
  model = "GCM_agg.txt"
  , monitor = poi
  , inits = init_values
  , data = size_data
  , n.chains = 3
  , sample = 5e4
  , burnin = 100
  , thin = 2
  , method = "rjparallel"
)
```

```{r echo = FALSE, results = "asis"}
knitr::kable(summary(size_samples)[c("c", "w", "b"), ])
```


<br />

# Unconstrained GCM fits for the angle condition
The original parameter estimates reported in Table 5 by Nosofsky (1989) are $w_1 = .98$, $c = 3.20$, and $b_1 = .43$.

```{r message = FALSE, results = "hide"}
memory <- as.matrix(subset(similarities, angle != 0)[, 1:2])
angle_data <- list(
  y = data$Cat.1.a
  , tests = tests
  , memory = memory
  , ntrials = data$n_angle
  , p = 2 # Shape of relationship between similarity and psychological distance
  , rho = 2 # Power of the Minkowski distance
  , category = unlist(subset(similarities, angle != 0)[, 4])
)

# Sample
angle_samples <- run.jags(
  model = "GCM_agg.txt"
  , monitor = poi
  , inits = init_values
  , data = angle_data
  , n.chains = 3
  , sample = 5e4
  , burnin = 100
  , thin = 2
  , method = "rjparallel"
)
```

```{r echo = FALSE, results = "asis"}
knitr::kable(summary(angle_samples)[c("c", "w", "b"), ])
```

<br />

# Predictions
The resulting fits allow for a close partial reproduction of Nosofsky's Figure 6 (1989) plotting observed against predicted proportions of category 1 responses for each stimulus.

```{r echo = FALSE, warning = FALSE}
par(pty = "s")
plot(NA, NA
  , xlim = c(0, 1)
  , ylim = c(0, 1)
  , xlab = "Observed probability"
  , ylab = "Predicted probability"
  , pch = 17
  , asp = 1
  , las = 1
)

all_size_samples <- coda::as.mcmc(size_samples)

for(i in 1:nrow(tests)) {
  vioplot(
    all_size_samples[, paste0("pred_y[", i, "]")] / data$n_size[i]
    , at = (data$Cat.1.s  / data$n_size)[i]
    , col = scales::alpha(grey(0.7), 0.35)
    , border = FALSE
    , rectCol = grey(0.5)
    , colMed = "black"
    , pchMed = 17
    , add = TRUE
    , wex = 0.15
  )
}

all_angle_samples <- coda::as.mcmc(angle_samples)

for(i in 1:nrow(tests)) {
  vioplot(
    all_angle_samples[, paste0("pred_y[", i, "]")] / data$n_angle[i]
    , at = (data$Cat.1.a  / data$n_angle)[i]
    , col = scales::alpha(grey(0.7), 0.35)
    , border = FALSE
    , rectCol = grey(0.5)
    , colMed = "black"
    , pchMed = 0
    , add = TRUE
    , wex = 0.15
  )
}

abline(0, 1)
legend(
  "topleft"
  , legend = c("Size", "Angle")
  , pch = c(17, 0)
  , inset = 0.1
  , bty = "n"
)
```

<br />

# References
Nosofsky, R. M. (1989). Further tests of an exemplar-similarity approach to relating identification and categorization. *Perception & Psychophysics*, 45(4), 279–290. doi: [10.3758/BF03204942](https://doi.org/10.3758/BF03204942)
