---
title: "Reproduction of model-based analyses by Nosofsky & Shin (1992)"
author: "Frederik Aust"
output:
  knitrBootstrap::bootstrap_document:
    highlight: xcode
    theme: flatly
    menu: false
---

To validate this implementation of the Generalized Context Model (GCM) for recognition data, I reproduced small parts of the model-based analyses reported in Nosofsky & Shin (1992). The original MDS solutions and response data were provided by Shin & Nosofsky (1992; s. [note on data](../data/README.html)).

```{r echo = FALSE, message = FALSE}
library("runjags")
library("vioplot")
```

# Experiment 1

```{r}
trials <- 3 * 50

tests <- c()
data <- c()
for(i in 1:3) {
  tests <- rbind(tests, read.csv2(paste0("../data/shin_nosofsky_1992_cat", i, ".csv")))
  data <- rbind(data, read.csv2(paste0("../data/shin_nosofsky_1992_responses_cat", i, ".csv")))
}
data$response <- round(data$Observed * trials)
```

<br />

## Summary fits

```{r message = FALSE, results = "hide"}
recognition_data <- list(
  y = data$response
  , tests = as.matrix(tests[, -1])
  , memory = as.matrix(subset(tests, Exemplar %in% paste0("O", 1:6))[, -1])
  , ntrials = rep(trials, nrow(tests))
  , p = 1 # Shape of relationship between similarity and psychological distance
  , rho = 2 # Power of the Minkowski distance
)

init_values <- list(
  list(c = 4, w_phi = c(-0.4, -2.2, -1, 1.3, -0.2, 1.1), k = 0.5)
  , list(c = 2, w_phi = c(1.4, 1.9, -0.1, -0.7, 0.8, 1.9), k = 0.3)
  , list(c = 3, w_phi = c(1, -0.4, -1.5, -0.6, 0.8, -1.8), k = 0.7)
)

poi <- c("c", "w", "k", "pred_y")

# Sample
gcm_samples <- run.jags(
  model = "GCM_agg_recognition2.txt"
  , monitor = poi
  , inits = init_values
  , data = recognition_data
  , n.chains = 3
  , sample = 5e4
  , burnin = 100
  , thin = 5
  , method = "rjparallel"
)
```

In Table 5 Shin & Nosofsky (1992) report the following estimates for the summary fits of the old-new recognition data in experiment 1: $w_1 = .006$, $w_2 = .084$, $w_3 = .102$, $w_4 = .392$, $w_5 = .218$, $c = 4.905$, $k = 0.280$

```{r echo = FALSE, results = "asis"}
knitr::kable(summary(gcm_samples)[c("c", paste0("w[", 1:6, "]"), "k"), ])
```

The resulting fits closely resemble those reported in the paper.

<br />

## Predictions
The resulting estimates allow for a close partial reproduction of Shin & Nosofsky's Figure 2A (1992) plotting observed against predicted proportions of old responses for each stimulus.

```{r echo = FALSE, warning = FALSE}
all_gcm_samples <- coda::as.mcmc(gcm_samples)

par(pty = "s")
plot(NA, NA
  , xlim = c(0, 1)
  , ylim = c(0, 1)
  , xlab = "Predicted recognition probability"
  , ylab = "Observed probability"
  , asp = 1
  , las = 1
)

for(i in 1:nrow(tests)) {
  vioplot(
    all_gcm_samples[, paste0("pred_y[", i, "]")] / trials
    , at = (data$response  / trials)[i]
    , horizontal = TRUE
    , col = scales::alpha(grey(0.7), 0.15)
    , border = FALSE
    , rectCol = grey(0.5)
    , colMed = "black"
    , pchMed = rep(c(1, rep(2, 6), rep(0, 3)), 3)[i]
    , add = TRUE
    , wex = 0.15
  )
}

abline(0, 1)
legend(
  "bottomright"
  , legend = c("Prototype", "Old", "New")
  , pch = c(1, 2, 0)
  , inset = 0.1
  , bty = "n"
)
```

<br />

# References
Shin, H. J., & Nosofsky, R. M. (1992). Similarity-scaling studies of dot-pattern classification and recognition. *Journal of Experimental Psychology: General*, 121(3), 278–304. doi:[10.1037/0096-3445.121.3.278](https://doi.org/10.1037/0096-3445.121.3.278)
