---
title: "Reproduction of model-based analyses by Nosofsky & Shin (1992)"
author: "Frederik Aust"
date: "19.12.2014"
output:
  html_document:
    theme: spacelab
    toc: yes
---

To validate this implementation of the Generalized Context Model (GCM) for recognition data, I reproduced small parts of the model-based analyses reported in Nosofsky & Shin (1992). The original MDS solutions and response data were provided by Shin & Nosofsky (1992; s. [note on data](data/README.html)).

```{r echo = FALSE}
source("gcm_rec_pred.r")
source("gcm_rec_fit.r")
```

# Experiment 1

```{r}
trials <- 3 * 50

obs <- c()
data <- c()
for(i in 1:3) {
  obs <- rbind(obs, read.csv2(paste0("data/shin_nosofsky_1992_cat", i, ".csv")))
  data <- rbind(data, read.csv2(paste0("data/shin_nosofsky_1992_responses_cat", i, ".csv")))
}
data$response <- round(data$Observed * trials)
```

<br />

## Summary fits

```{r}
mem <- subset(obs, Exemplar %in% paste0("O", 1:6))

ui <- structure(c(1, -1, 1, 0, 0, 0, 0, 0, 0, 1, -1, 0, 1, 0, 0, 0, 0, 0, 1, -1, 0, 0, 1, 0, 0, 0, 0, 1, -1, 0, 0, 0, 1, 0, 0, 0, 1, -1, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1), .Dim = c(9L, 7L), .Dimnames = list(NULL, NULL))
ci <- c(0, -1, 0, 0, 0, 0, 0, 0, 0)

gcm_fit <- constrOptim(
  theta = c(w1 = 1/6, w2 = 1/6, w3 = 1/6, w4 = 1/6, w5 = 1/6, c = 3, k = 1)
  , f = gcm_rec_fit
  , mem = mem[, -1]
  , obs = obs[, -1]
  , rho = 2
  , n = trials
  , data = data
  , method = "Nelder-Mead"
  , ui = ui
  , ci = ci
)
```

In Table 5 Shin & Nosofsky (1992) report the following estimates for the summary fits of the old-new recognition data in experiment 1: $w_1 = .006$, $w_2 = .084$, $w_3 = .102$, $w_4 = .392$, $w_5 = .218$, $c = 4.905$, $k = 0.280$

```{r}
round(gcm_fit$par, 3)
gcm_fit$value/2 # -lnL
```

The resulting fits closely resemble those reported in the paper, however, the estimates vary slightly depending on the starting parameters `theta` used when fitting the data.

<br />

## Predictions
The resulting estimates allow for a close partial reproduction of Shin & Nosofsky's Figure 2A (1992) plotting observed against predicted proportions of old responses for each stimulus.

```{r echo = FALSE}
cat_pred <- gcm_rec_pred(gcm_fit$par, mem[, -1], obs[, -1])

par(pty = "s")
plot(
  cat_pred
  , data$Observed
  , xlab = "Predicted recognition probability"
  , ylab = "Observed probability"
  , xlim = c(0, 1)
  , ylim = c(0, 1)
  , pch = c(1, rep(2, 6), rep(0, 3))
  , asp = 1
  , las = 1
)
abline(0, 1)
legend(
  "bottomright"
  , legend = c("Prototype", "Old", "New")
  , pch = c(1, 2, 0)
  , inset = 0.1
  , bty = "n"
)
```

<br />

# References
Shin, H. J., & Nosofsky, R. M. (1992). Similarity-scaling studies of dot-pattern classification and recognition. *Journal of Experimental Psychology: General*, 121(3), 278–304. doi:[10.1037/0096-3445.121.3.278](https://doi.org/10.1037/0096-3445.121.3.278)
